---
title: "Geoscience of the Future"
author: "Fulweiler, Emery, Maguire"
date: "Thursday, May 14, 2015"
output: html_document
---

Starting with Fig 3, be sure to set your working directory to the Dropbox shared folder for now. If you are using a Mac, the directory will be set automatically, but you'll have to do it manually if you are using a PC.

```{r,echo=FALSE}
#### this is me making a change!!! ####

if(.Platform$OS.type=="unix"){
  setwd("~/Dropbox/GFP_N2NarragansettBay")
}

# for Tim / PC use
# setwd("C:/Users/Tim/Dropbox/GFP_N2NarragansettBay")

library(plotrix)
library(Hmisc)
library(plyr)

se<-function(x) sqrt(var(x,na.rm=TRUE)/length(na.omit(x)))

```


#Figure 3

Notes: the error bars (standard error) in the Excel file seem to be MUCH bigger than what's in the paper for Prov. The Bay error bars look OK.

```{r,fig.width=10,echo=FALSE}
#### data import ####
mimsdata <- read.csv("Fig4.csv", stringsAsFactors=FALSE)
mimsdata = mimsdata[c(-7,-8,-9),] # Prov 4/30/2006 look like doubles from Mid Bay, by dumping these the linear model in Figure 4 has the same equations as the paper
mimsdata$date=format(as.POSIXct(strptime(mimsdata$Date,format="%m/%d/%y",tz="EST")),"%Y-%m")
mimsdata = na.omit(mimsdata)

providence = cbind(adply(tapply(mimsdata[mimsdata$Site=="Prov",]$N2Flux,mimsdata[mimsdata$Site=="Prov",]$date,mean), c(1)),adply(tapply(mimsdata[mimsdata$Site=="Prov",]$N2Flux,mimsdata[mimsdata$Site=="Prov",]$date,se), c(1)))[,-3]
colnames(providence) = c("date","ProvN2_mean","ProvN2_stderr")

midbay = cbind(adply(tapply(mimsdata[mimsdata$Site=="Mid Bay",]$N2Flux,mimsdata[mimsdata$Site=="Mid Bay",]$date,mean), c(1)),adply(tapply(mimsdata[mimsdata$Site=="Mid Bay",]$N2Flux,mimsdata[mimsdata$Site=="Mid Bay",]$date,se), c(1)))[-10,-3] # there is a one-off January 2010 Mid Bay sample?
colnames(midbay) = c("date","BayN2_mean","BayN2_stderr")

dat = merge(providence,midbay,by="date", all=T)
dat$date = as.Date(paste(dat$date,"-01",sep=""))
dat = dat[order(dat$date),]
dat[,2] = round(dat[,2],2)
dat[,3] = round(dat[,3],2)
dat[,4] = round(dat[,4],2)
dat[,5] = round(dat[,5],2)
dat$date=format(as.POSIXct(strptime(dat$date,format="%Y-%m-%d",tz="EST")),"%b-%y")


par(mfrow = c(1,2))
#plot A - Providence River Estuary
par(mai=c(1.4,1.4,1,0.1))
mean <- dat$ProvN2_mean
sderr <- dat$ProvN2_stderr
plotCI(barplot(mean,
               col="green4",
               # y axis:
               ylim=c(-300,300), 
               cex.axis=.7,
               ylab=expression(Net ~ Sediment ~ N[2]-N ~ Flux ~ (mu ~ mol ~ m^{-2} ~ hr^{-1})),
               cex.lab=.7,
               tcl=-.2,
               #x axis:
               names=dat$date, 
               las=2, #makes labels perpendicular to axis
               cex.names=.7
               ),
       mean,
       sderr,
       pch=NA, #suppresses center point on error bars
       sfrac=0, #suppresses serifs on error bars
       add=TRUE)
abline(h=0)
abline(h=mean(mean,na.rm=TRUE),lty=2,lwd=1.5)
text(1,250,"a")

#plot B - Narragansett Bay
par(mai=c(1.4,0.1,1,1.4))
mean <- dat$BayN2_mean
sderr <- dat$BayN2_stderr
plotCI(barplot(mean,
               col="red3",
               ylim=c(-300,300),  
               yaxt="n", #supresses y axis
               names=dat$date, 
               las=2, #makes labels perpendicular to axis
               cex.names=.7),
       mean,
       sderr,
       pch=NA, #suppresses center point on error bars
       sfrac=0, #suppresses serifs on error bars
       add=TRUE)
axis(2,labels=FALSE,tcl=-.2) #adds an axis without labels
abline(h=0)
abline(h=mean(mean,na.rm=TRUE),lty=2,lwd=1.5)
text(1,250,"b")

par(mfrow = c(1,1))
```

Figure 3. Monthly net sediment N$_2$-N flux (± standard error) across the sediment-water interface over a nine-year record for (a) the Providence River Estuary, and (b) the historic mid-Narragansett Bay site. Overall nine-year mean N$_2$-N flux is highlighted in each (dashed line).

# Figure 4



```{r,fig.width=8, echo=FALSE}
par(mai=c(1.02,0.82,0.82,0.42))

Fig4 <- mimsdata

Fig4$sym = 0
Fig4[which(Fig4[,1] == unique(Fig4[,1])[1]),]$sym = 19
Fig4[which(Fig4[,1] == unique(Fig4[,1])[2]),]$sym = 1

lmo2 = Fig4$O2Flux/2
N2_fit = lm(Fig4$N2Flux~lmo2)
#summary(N2_fit)
#range(Fig4[,5])

plot(Fig4$O2Flux/2,Fig4$N2Flux, pch=Fig4$sym,  ylim = c(-300,350),xlab="",ylab="")
abline(N2_fit)
mtext(side = 2, text = expression(Net ~ Sediment ~ N[2]-N ~ Flux ~ (mu ~ mol ~ m^{-2} ~ h^{-1})), line = 1.8)
mtext(side = 1, text = expression(O[2] ~ Flux ~ (mu ~ mol ~ m^{-2} ~ h^{-1})), line = 2.5)



```

Figure 4. Individual net sediment N$_2$ fluxes as a function of sediment oxygen demand for the Providence River Estuary (closed circles) and mid-Narragansett Bay (open circles). N$_2$ was significantly correlated (0.0006) to sediment oxygen demand but the overall predictive power of the relationship was weak (N$_2$-N flux = 0.038(O$_2$ flux) – 5.58; R$^2$ = 0.09, n = 134).

# Figure 5
```{r,fig.width=8,echo=FALSE}

Fig5 <- read.csv("Fig5.csv", stringsAsFactors=FALSE)
#Fig5 <- na.omit(Fig5)

exp.lm <- lm(log(Fig5$chl_a)~Fig5$Year)
#summary(exp.lm)
#exp(exp.lm$coeff[1])
#round(exp.lm$coeff[2],3)

chl.exponential <- exp(predict(exp.lm,list(Year=Fig5$Year)))

plot(Fig5[,1],
     Fig5[,2],
     xlab="",
     ylab="",
     pch = 21,
     bg="green4",
     ylim = c(0,12),
     yaxs="i",
     xaxt="n")
axis(1, at=seq(1970,2015,5), labels=seq(1970,2015,5))
minor.tick(ny=0,nx=10, tick.ratio=.5)
mtext(side = 2, text = expression(Chlorophyll ~ italic(a) ~ (mg ~ m^{-3})), line = 1.8)
mtext(side = 1, text = expression(Year), line = 2.5)
lines(Fig5$Year, chl.exponential,lwd=2)

```

Figure 5. Mean annual water column chlorophyll a concentrations over the past four decades at the long-term monitoring station in the middle of Narragansett Bay’s west passage, expanded from Nixon et al. (2009) (Chl *a* = 3.0 x 10$^{20}$ e$^{-??0.023(Year)}$; R$^2$ = .40). The mean annual chlorophyll decline is significant (p < .0001). Annual data from Li and Smayda (1998) through 1990. Historical data (1991-??1997) from Ted Smayda (University of Rhode Island) based on personal communication from Scott Nixon). All data from 1999 to present are from the University of Rhode Island Graduate School of Oceanography (GSO) plankton monitoring program (http://www.gso.uri.edu/phytoplankton); the year 2012 is not included as there were only two months of chlorophyll data available. Note that in May 2007, the GSO monitoring program changed to an immediate extraction method. The new and traditional extraction techniques were compared over a one-year period (July 2007 to July 2008; Graff and Rynearson, 2011). We used this year-long study to correct data from May 2007 to the present so that we could compare the long-term trend in chlorophyll. (All new data were multiplied by the slope of the linear regression of the traditional method as a function of the new method, y = 0.4323x -?? 1.2094; R$^2$ = 0.88).

# Figure 6

Notes: It looks like the triangle symbols (filled vs open) were switched in the paper. Should I make it look like the paper, or make it do as the legend says?   
Also, the figure legend in the paper says 1985, but the data says 1986.   
The regression lines could be cleaned up if we used segments() instead of abline(), but then it would be harder to plug the fitted equation in.


```{r,fig.height=7,fig.width=10,echo=FALSE}
# data import
dat=read.csv("Fig6.csv",header=TRUE)

#fix years
dat$yr=as.numeric(as.character(dat$Year))
dat$yr[is.na(dat$yr)]=1981

# split data into current and historical groups 
recent=dat[dat$yr>=2005,]
z1979 <- dat[dat$yr==1979,]
MERL <- dat[dat$yr==1981,]
z1986 <- dat[dat$yr==1986,]

# three fits described in table 1
allfit=lm(N2_flux~chl_a,data=dat) #all the data
netfit=lm(N2_flux~chl_a,data=recent) #just the new data
denitfit=lm(N2_flux~chl_a,data=recent[recent$N2_flux>0,]) #just the new POSITIVE data

#plot the current data 
par(mai=c(1.4,1.4,1,1.4))
plot(recent$chl_a,
     recent$N2_flux,
     pch=21,
     bg="grey",
     xlim=c(0,60),
     ylim=c(-200,1000),
     ylab="",
     xlab="",
     xaxt="n", #suppress the axis
     yaxt="n",
     bty="n") # suppress the box
axis(1, pos=0) # add the x-axis at y=0
mtext(expression(Chlorophyll ~ italic(a) ~ (mg ~ m^{-3})),1,-1)
axis(2, pos=0)
mtext(expression(Net ~ Sediment ~ N[2]-N ~ Flux ~ (mu ~ mol ~ m^{-2} ~ hr^{-1})),2,2)
abline(allfit) # add the regression line from the complete data
points(MERL$chl_a,MERL$N2_flux,pch=22) # add the MERL points as open squares
points(z1979$chl_a,z1979$N2_flux,pch=17) # add the 1979 points as closed triangles
points(z1986$chl_a,z1986$N2_flux,pch=2) # add the 1986 points as open triangles

#set bounds in par for the inset figure
par(fig=c(0.55,1,.2,.8),new=TRUE)

#plot the inset - just recent data
plot(recent$chl_a,
     recent$N2_flux,
     pch=21,
     bg="grey",
     xlim=c(0,5),
     ylim=c(-200,150),
     mgp=c(2,1,0),
     cex.axis=.8,
     cex.lab=.8,
     ylab=expression((mu ~ mol ~ m^{-2} ~ hr^{-1})),
     xlab="",
     xaxt="n")
mtext(expression(Net ~ Sediment ~ N[2]-N ~ Flux),2,3,cex=.8)
mtext(expression(Chl ~ italic(a) ~ (mg ~ m^{-3})),1,0.5,cex=.8)
axis(1,pos=0,cex.axis=.8)
abline(h=0)
abline(netfit) # add the fit to this data set
par(fig=c(0,1,0,1)) #reset the bounds to the default
```

Figure 6. Mean summer (June, July, August) net sediment N$_2$-N flux as a function of summer surface water column chlorophyll a for sediments from mid-Narragansett Bay over the last 40 years. Shaded circles = 2005–2013. Closed triangle = 1979 (Seitzinger et al., 1984). Open triangle = 1985 (Nowicki, 1994). Open squares = MERL 1981 (Seitzinger, 1982). Unpublished summer chlorophyll a data from 1968 to 1996 is from Ted Smayda (University of Rhode Island). This relationship is significant (p < 0.0001). The inset shows the relationship between mean summer water column chlorophyll and net N$_2$ fluxes from 2005 to 2013 only (p = 0.0505). See Table 1 and the text for more details.

# Table 1

Table 1. Equations used to predict mean summer sediment N$_2$ fluxes using the relationships between mean summer water column surface chlorophyll and directly measured sediment N$_2$ fluxes.

Description | Equation | $R^2$ | p-value | n   
----|----|----|----|----   
$^1$All N$_2$ Flux Data | y = `r round(allfit$coefficients[2],1)` * chl  + `r round(allfit$coefficients[1],1)` | `r round(summary(allfit)$r.squared,2)` | `r summary(allfit)$coefficients[2,4]` | `r dim(dat)[1]`   
$^2$Net N$_2$ Fluxes | y = `r round(netfit$coefficients[2],1)` * chl  + `r round(netfit$coefficients[1],1)` | `r round(summary(netfit)$r.squared,2)` | `r summary(netfit)$coefficients[2,4]` | `r dim(recent)[1]`
$^3$Net Denitrification Fluxes | y = `r round(denitfit$coefficients[2],1)` * chl  + `r round(denitfit$coefficients[1],1)` | `r round(summary(denitfit)$r.squared,2)` | `r summary(denitfit)$coefficients[2,4]` | `r dim(recent[recent$N2_flux>0,])[1]`

$^1$ This equation includes all historic and recent N$_2$ flux measurements including those from the MERL experiments.    
$^2$ This equation only uses N$_2$ fluxes from 2005–2013 measured with the N$_2$/Ar technique.   
$^3$ This equation only uses net positive N$_2$ fluxes (i.e., net denitrification) from 2005–2013 measured with the N$_2$/Ar technique.

# Figure 7
```{r, fig.width=8,echo=FALSE}
# data import
dat=read.csv("Fig7.csv",header=TRUE)

# plot
mean=rbind(dat$N2fit_mean,dat$N2flux_mean)
sderr=rbind(dat$N2fit_stderr,dat$N2flux_stderr)

par(mai=c(1.4,1.4,1,.2))
plotCI(barplot(mean,
               col=c("steelblue4","maroon"),
               ylim=c(-200,1000),
               cex.axis=.7,
               ylab=expression(Net ~ Sediment ~ N[2]-N ~ Flux ~ (mu ~ mol ~ m^{-2} ~ hr^{-1})),
               cex.lab=.7,
               tcl=-.2,
               
               names=dat$Year,
               las=2, #makes labels perpendicular to axis
               cex.names=.7,
               #xaxt="n", #supresses x axis
               space=c(0,0),
               beside=TRUE, # as opposed to stacked
               border=NA
  ),
  mean,
  sderr,
  pch=NA, #suppresses center point on error bars
  sfrac=0, #suppresses serifs on error bars
  add=TRUE)
abline(h=0)
```

Figure 7. Long-term mean (± standard error) summer net sediment N$_2$ flux rates in Narragansett Bay. Blue bars = predicted rates. Red bars = measured rates. The predicted rates shown here are the average of the rates predicted from the three different relationships between mean summer (June, July, August) surface water column chlorophyll and measured sediment N$_2$ fluxes (Table 1: All, 2005–2013 all, and 2005–2013 net denitrification only) (± standard error of the three different predictions).
